---
title: "Joining Data"
author: "Murray Cadzow"
date: "`r lubridate::today()`"
output:
    html_document: 
      theme: cerulean
      highlight: haddock
      toc: yes
      toc_float: yes
      number_sections: yes
      code_folding: "show"
#bibliography: references.bib
---

```{r setup, message = FALSE, warning=FALSE}
library(tidyverse)
library(here)
library(palmerpenguins)
```

# Bringing data together

When we're talking about data, often we'll have a situation where we:

a) Have the same columns and want to combine data sets together
b) Have multiple data sets with differing columns that have an unique column(s) in common that we want to bring data together from.



## Binds

### Adding rows

This is a very common thing you would want to do. A situation that often leads to wanting to do this is having your data split into separate files based on a category (e.g. location) and you want to read the data in and combine it into a single data set.

Key to this is that you have numerous columns in common between all the data sets.

In base R this is an `rbind()` (row bind) where you need to have the same columns and want to combine all the rows together. 

In `dplyr` this is `bind_rows()` and it will add the rows where they match on column names but where they don't it will fill in `NA`s where the data didn't previously exist.

The general idea is that the number of rows is changing but the number of columns is the union of the column names from all sources.

With the penguin data that has a file per island it could look like this:
```{r, message=FALSE}
biscoe <- read_csv(here("Data/biscoe.csv"))
names(biscoe)
dim(biscoe)
dream <- read_csv(here("Data/dream.csv"))
names(dream)
dim(dream)



row_combined <- bind_rows(biscoe, dream)
names(row_combined)
dim(row_combined)
```

 (remember the last session at https://github.com/murraycadzow/R_functions_and_iterators for an example of scaling this idea to lots of files)

### Adding columns

Adding new columns to a data set is also something that we want to be able to do. Often this is because we have additional data about the samples we have that we want to 'add onto the side'. In this case the number of columns changes but the number of rows stays the same.

In base R this is a `cbind()`. In `dplyr` this is `bind_cols()`.

```{r}
biscoe_pt1 <- biscoe %>% select(1:5)

biscoe_pt2 <- biscoe %>% select(6:9)

column_combined <- bind_cols(biscoe_pt1, biscoe_pt2)
dim(column_combined)
```



**An important caveat is the row order need to correspond exactly between datasets.** A join/merge is a safer option for this reason.


## Joins

Joins (also called merges) are a much safer option than just slapping columns onto the side of your data with `cbind()` or `bind_cols()`. The reason for this is that joins rely on the use of a _key_ which is a column(s) that contains a unique value per row (often this would be an id column) and is in common between the data sets. What makes this safer is that R will match the rows together based on this identifier rather than you having to ensure the data is in exactly the same row order.

The `dplyr` cheat sheets have some really useful diagrams to show what is going on for each type of join.

### Inner join

An inner join does the intersection of the _key_ columns. If the key value is found in both data sets the row is kept, if the key value is only in one data set the row is dropped.

Consider

```{r}
(df1 <- tibble(id = 1:3, y = letters[1:3]))
(df2 <- tibble(id = 2:4, z = c("word1", "word2", "word3")))

# keeps only rows with id matching in both data sets
inner_join(df1, df2, by = "id")
```

### Left/Right Join

Left or right join will keep all of the rows from the data set that is specified on the left for `left_join()`, or right for `right_join()`. These are often used when you have a starting data set and you want to add on extra information where it exists but you're not worried about completeness of all your rows.

```{r}
# keeps all rows in df1
left_join(df1, df2, by = "id")

# keeps all rows in df2
right_join(df1, df2, by = "id")
```

### Full Join

Full join will keep all rows from both data sets 

```{r}
# keeps all rows from both data sets
full_join(df1, df2, by = "id")
```

### Duplicated keys

Some important things to remember is that with joins, if you have duplicated keys, then you end up with extra combinations because the join does a cross-product (creates every possible combination).


Consider these two data sets
```{r}
(df_a <- tibble(animal = c("dog","cat","fish", "cat"), name = c("woofy", "pawsy", "bubbles", "meows")))
(df_b <- tibble(animal = c("dog", "cat","fish", "cat"), house = c("kennel", "bed", "tank", "blankey")))
```

and we want to create
```{r, echo = FALSE}
bind_cols(df_a, df_b[,-1]) 
```


```{r}
left_join(df_a, df_b, by = "animal")

inner_join(df_a, df_b, by = "animal")

full_join(df_a, df_b, by = "animal")
```
We could column bind this data, or we could (safer) create a unique key and then join.


## Example

We want to add some additional information onto our Penguin data. In this case lets add the latitude and longitude of the islands.

```{r}
island_coords <- tribble(
  ~island, ~latitude, ~longitude,
"Dream", -64.7333323,-64.2420877,
"Torgersen", -64.7666663, -64.0920878,
"Biscoe", -66.0770889,-66.8244735, 
)
```


Now we can add this information onto our data set `palmerpenguins::penguins`
```{r, include = FALSE}
left_join(penguins, island_coords, by = "island")
```




